Detection of Oscillations in Aldebaran with Ground-Based Observations

### 2018-01-17 paper/aldebaran.tex
WIP on archival data section, time series analysis section.

A Gaussian Process such as that used by \citet{brewer2009} also
corresponds to the solution, $y(t)$, of the stochastic ODE
\begin{equation}
  \label{eq:GP-definition}
  \prod_{i=1}^{p} \left[ \diff{}{t} - r_i \right] y(t) = \prod_{j = 1}^{q} \left[ \diff{}{t} - b_i \right] \eta(t),
\end{equation}
where $p, q \in \mathbb{Z}$, $p > q \geq 0$ define the order of the
process; $r_i \in \mathbb{C}$, $\Re r_i < 0$ are the roots of the
characteristic equation of the ODE; $b_j \in \mathbb{C}$,
$\Re b_j < 0$ are corresponding roots of the inverse process; and
$\eta(t)$ is a white-noise Gaussian Process with
$\left\langle \eta(t) \right\rangle = 0$ and
$\left\langle \eta(t) \eta(t') \right\rangle = \sigma^2 \delta\left(t
  - t' \right)$.  Such a solution is called a CARMA (Continuous
AutoRegressive Moving Average) process.  The constraint that
$\Re r_i < 0$ and $\Re b_j < 0$ ensures that the linear operator
defining the process and its inverse are invertable.  Under the
assumption that the solution, $y(t)$, is real, roots $r_i$ and $b_j$
must either be real or occur in complex conjugate pairs.  The power
spectrum of the process is
\begin{equation}
  \label{eq:CARMA-PSD}
  P_y(f) = \sigma^2 \left| \frac{\prod_{j=1}^q \left[ 2 \pi i f - b_j \right]}{\prod_{i=1}^p \left[ 2 \pi i f - r_i \right]} \right|^2.
\end{equation}
The power spectrum is a rational function of frequency with poles at
$2\pi i f = r_i$ and zeros at $2\pi i f = b_i$; because of the
invertable constraints, these poles and zeros all occur when
$\Im f \neq 0$.  The autocorrelation function corresponding to the
power spectrum in Eq.\ \eqref{eq:CARMA-PSD} is
\begin{equation}
  \label{eq:CARMA-ACF}
  \rho_y (\tau) = \left\langle y\left( t \right) y \left( t + \tau \right) \right\rangle = \sum_{i=1}^p A_i^2 e^{r_i \tau},
\end{equation}
where the RMS amplitudes of the $p$ different modes, $A_i$, are
functions of $\sigma$ and the values of the roots $r_i$ and $b_j$.  If
$q = p-1$, then the $A_i$ are independent, while if $q < p-1$ then
there are interdependencies among the $A_i$.  From either the power
spectrum in Eq.\ \eqref{eq:CARMA-PSD} or the autocorrelation function
in Eq.\ \eqref{eq:CARMA-ACF}, it is apparent that a real root $r_i$
represents an exponentially-decaying mode with $r_i$ the rate
constant, while a complex conjugate pair of $r_i$ represent an
oscillatory mode with decay rate $\Re r_i$ and angular frequency
$\left|\Im r_i\right|$.  The former is called an CAR(1) process, an
Ornstein-Uhlenbeck process, or a damped random walk.
\citet{Kelly2014} gives more details on CARMA processes, including
formulae for $A_i$ as a function of $\sigma$, $r_i$, and $b_j$, and
shows how a state-space model of the solution $y(t)$ can be tracked
through a time-series of uncertain observations
$y_k = y\left( t_k \right) + \epsilon_k$, with Gaussian noise
$\epsilon_k$ of zero mean and known (heteroskedastic) variance, using
a Kalman filter to produce a likelihood function
$p\left( y_k \mid \sigma, r_i, b_j \right)$ in
$\mathcal{O}\left( n \right)$ computational time.  Additionally,
\citet{Foreman-Mackey2017} show how a dimensional expansion can be
used to reduce the standard Gaussian Process covariance matrix for the
process with autocorrelation given by Eq.\ \eqref{eq:CARMA-ACF} under
observations at times $t_k$ to a banded form that can then be
Cholesky-decomposed in the standard Gaussian Process likelihood
function in $\mathcal{O}\left( n \right)$ time.

Here we have adopted the \citet{Kelly2014} approach.  The \texttt{CARMA.jl} package\footnote{\url{https://github.com/farr/CARMA.jl}} implements Kalman filters that can compute the likelihood for a set of observations parameterised by either the roots $r_i$ and $b_j$ or the RMS amplitudes $A_i$ and roots $r_i$.  Because we have a set of observations taken with different instruments at different sites (see Section \ref{sec:archival-obs}), we include RV offset (i.e.\ mean value) parameters that are site-dependent and also a site-dependent scaling parameter.  Thus, the $i$th RV measured at site $j$, $y_{ij}$, is assumed to be
\begin{equation}
\label{eq:RV-model}
y_{ij} = v_r\left(t_i\right) + \mu_j + y\left( t_i \right) + \nu_j \epsilon_{ij},
\end{equation}
$t_i$ is the time of the measurement, $v_r$ is a planetary RV signal \citep{Hatzes2015,Gregory2005}, $\mu_j$ is the site-dependent mean RV offset, $y(t)$ is the CARMA process representing stellar activity, $\nu_j$ is the site-dependent uncertainty scaling parameter, and $\epsilon_{ij}$ is an independent Gaussian random variable with mean zero and standard deviation equal to the quoted uncertainty of the observation.  We impose broad priors (see Table \ref{tab:parameters}) on the parameters of the planetary RV signal, the CARMA process, the $\mu_j$ offsets, and the $\nu_j$ scale factors.

\begin{deluxetable}{lll}
\tablecolumns{3}
\tablecaption{\label{tab:parameters} The parameters in our model (see Eq.\ \eqref{eq:RV-model}).}
\tablehead{\colhead{Parameter} & \colhead{Prior} & \colhead{Description}}
\startdata
\cutinhead{Per-Site RV Offset and Uncertainty Scaling}
$\mu_j$ & $N\left[ \left\langle y_{j} \right\rangle, 10 \sigma_{y_{j}} \right] \left( \mu _j \right)\tablenotemark{a}$ & The per-site RV offset. \\
$\nu_j$ & $1/\nu_j, \quad \nu_j \in [1/2, 2]$ & The per-site uncertainty scale factor. \\
\cutinhead{Keplerian Parameters}
$K$ & $1/K, \quad K \in \left[ \min_j \sigma_{y_j}/100, \max_j 10 \sigma_{y_j} \right]$ & RV semi-amplitude. \\
$P$ & $1/P, \quad P \in [600, 700] \, \mathrm{d}$ & Keplerian period \\
$e$ & $2e,\quad 0 \leq e < 1$\tablenotemark{b} & Keplerian eccentricity \\
$\omega$ & $\mathrm{const}, \quad 0 \leq \omega < 2\pi$ & Longitude of pericentre \\
$\chi$ & $\mathrm{const}, \quad 0 \leq \chi < 1$ & Pericentre passage is at $t = \chi P$\\
\enddata
\tablenotetext{a}{$\left\langle y_j \right\rangle$ and $\sigma_{y_j}$ are the mean and standard deviation of the RV timeseries from site $j$.}
\tablenotetext{b}{We actually sample in and impose a (flat) prior on $x \equiv e \cos \omega$ and $y \equiv e \sin \omega$.}
\end{deluxetable}

\section{Aldebaran}
Aldebaran is a red giant star with spectral type K5, one of the nearest such stars at a distance of only $19.96 \pm 0.38$~pc as determined by \hipparcos \citep{hipparcos}. Its position near the Ecliptic permits the determination of its angular diameter by lunar occultations and by interferometry \citep[$20.58 \pm 0.03$ mas;][]{richichi2005,1979ApJ...228L.111B,brown1979,panek1980}. These tight constraints are valuable in breaking degeneracies in stellar modelling and make this an ideal star for asteroseismic characterization (Appendix~/ref{sec:mod}).

@@

>>>>

A Gaussian Process such as that used by \citet{brewer2009} also corresponds to
the solution, $y(t)$, of the stochastic ODE \begin{equation}
\label{eq:GP-definition} \prod_{i=1}^{p} \left[ \diff{}{t} - r_i \right] y(t) =
\prod_{j = 1}^{q} \left[ \diff{}{t} - b_i \right] \eta(t), \end{equation} where
$p, q \in \mathbb{Z}$, $p > q \geq 0$ define the order of the process; $r_i \in
\mathbb{C}$, $\Re r_i < 0$ are the roots of the characteristic equation of the
ODE; $b_j \in \mathbb{C}$, $\Re b_j < 0$ are corresponding roots of the inverse
process; and $\eta(t)$ is a white-noise Gaussian Process with $\left\langle
\eta(t) \right\rangle = 0$ and $\left\langle \eta(t) \eta(t') \right\rangle =
\sigma^2 \delta\left(t - t' \right)$.  Such a solution is called a CARMA
(Continuous AutoRegressive Moving Average) process.  The constraint that $\Re
r_i < 0$ and $\Re b_j < 0$ ensures that the linear operator defining the process
and its inverse are invertable.  Under the assumption that the solution, $y(t)$,
is real, roots $r_i$ and $b_j$ must either be real or occur in complex conjugate
pairs.  The power spectrum of the process is \begin{equation}
\label{eq:CARMA-PSD} P_y(f) = \sigma^2 \left| \frac{\prod_{j=1}^q \left[ 2 \pi i
f - b_j \right]}{\prod_{i=1}^p \left[ 2 \pi i f - r_i \right]} \right|^2.
\end{equation} The power spectrum is a rational function of frequency with poles
at $2\pi i f = r_i$ and zeros at $2\pi i f = b_i$; because of the invertable
constraints, these poles and zeros all occur when $\Im f \neq 0$.  The
autocorrelation function corresponding to the power spectrum in Eq.\
\eqref{eq:CARMA-PSD} is \begin{equation} \label{eq:CARMA-ACF} \rho_y (\tau) =
\left\langle y\left( t \right) y \left( t + \tau \right) \right\rangle =
\sum_{i=1}^p A_i^2 e^{r_i \tau}, \end{equation} where the RMS amplitudes of the
$p$ different modes, $A_i$, are functions of $\sigma$ and the values of the
roots $r_i$ and $b_j$.  If $q = p-1$, then the $A_i$ are independent, while if
$q < p-1$ then there are interdependencies among the $A_i$.  From either the
power spectrum in Eq.\ \eqref{eq:CARMA-PSD} or the autocorrelation function in
Eq.\ \eqref{eq:CARMA-ACF}, it is apparent that a real root $r_i$ represents an
exponentially-decaying mode with $r_i$ the rate constant, while a complex
conjugate pair of $r_i$ represent an oscillatory mode with decay rate $\Re r_i$
and angular frequency $\left|\Im r_i\right|$.  Oscillatory modes can also be
described by their frequency $f$ and quality factor $Q$ via $r = f/(2Q) \pm 2\pi
i f$.  Decaying modes generate a damped random walk, also known as a
Ornstein-Uhlenbeck process.  \citet{Kelly2014} gives more details on CARMA
processes in an astronomical context, including formulae for $A_i$ as a function
of $\sigma$, $r_i$, and $b_j$, and shows how a state-space model of the solution
$y(t)$ can be tracked through a time-series of uncertain observations $y_k =
y\left( t_k \right) + \epsilon_k$, with Gaussian noise $\epsilon_k$ of zero mean
and known (heteroskedastic) variance, using a Kalman filter to produce a
likelihood function $p\left( y_k \mid \sigma, r_i, b_j \right)$ in
$\mathcal{O}\left( n \right)$ computational time.  Additionally,
\citet{Foreman-Mackey2017} show how a dimensional expansion can be used to
reduce the standard Gaussian Process covariance matrix for the process with
autocorrelation given by Eq.\ \eqref{eq:CARMA-ACF} under observations at times
$t_k$ to a banded form that can then be Cholesky-decomposed in the standard
Gaussian Process likelihood function in $\mathcal{O}\left( n \right)$ time.

Here we have adopted the \citet{Kelly2014} approach.  The \texttt{CARMA.jl} package\footnote{\url{https://github.com/farr/CARMA.jl}} implements Kalman filters that can compute the likelihood for a set of observations parameterised by either the roots $r_i$ and $b_j$ or the RMS amplitudes $A_i$ and roots $r_i$.  We also include a Keplerian RV signal from a planetary companion to Aldebaran \citep{Hatzes2015,Gregory2005}.  Because we have a set of observations taken with different instruments at different sites (see Section \ref{sec:archival-obs}), we include RV offset (i.e.\ mean value) parameters that are instrument- and site-dependent and also an instrument- and site-dependent scaling parameter.  Thus, the $i$th RV measured at site and intstrument combination $j$, $y_{ij}$, is assumed to be
\begin{equation}
  \label{eq:RV-model}
  y_{ij} = v_r\left(t_i\right) + \mu_j + y\left( t_i \right) + \nu_j \epsilon_{ij},
\end{equation}
$t_i$ is the time of the measurement, $v_r$ is the planetary RV signal, $\mu_j$ is the mean RV offset for site and instrument combination $j$, $y(t)$ is the CARMA process representing stellar activity, $\nu_j$ is the uncertainty scaling parameter for site and instrument combination $j$, and $\epsilon_{ij}$ is an independent Gaussian random variable with mean zero and standard deviation equal to the quoted uncertainty of the observation.  We impose broad priors (see Table \ref{tab:parameters}) on the parameters of the planetary RV signal, the CARMA process, the $\mu_j$ offsets, and the $\nu_j$ scale factors.  We treat the frequency of maximum asteroseismic power,  $\nu_\mathrm{max}$, as the frequency of an appropriate oscillatory mode in the CARMA process representing the stellar activity; this is equivalent to fitting a Lorentzian profile to a set of unresolved asteroseismic modes.  We used the nested sampling algorithm of the \texttt{Ensemble.jl} package\footnote{\url{https://github.com/farr/Ensemble.jl}; this package implements several stochastic sampling algorithms based on the ``stretch move'' proposal used in \texttt{emcee} \citep{Foreman-Mackey2013} and introduced in \citet{Goodman2010}} to calculate the marginal likelihood (evidences) and draw samples from the posterior distribution over the parameters of our various models.

\begin{deluxetable}{lll}
\tablecolumns{3}
\tablecaption{\label{tab:parameters} The parameters in our model Eq.\ \eqref{eq:RV-model} and their priors.  Site and instrument combinations are indexed by $j$.  Modes in the stochastic process are indexed by $k \in [1,p]$ and are either real (parameterised by a rate parameter $r_k$) or complex (parameterised by a frequency $f_k$ and quality factor $Q_k$).}
\tablehead{\colhead{Parameter} & \colhead{Prior} & \colhead{Description}}
\startdata
\cutinhead{Site/Instrument RV Offset and Uncertainty Scaling}
$\mu_j$ & $N\left[ \left\langle y_{j} \right\rangle, 10 \sigma_{y_{j}} \right] \left( \mu _j \right)\tablenotemark{a}$ & The site/instrument RV offset. \\
$\nu_j$ & $1/\nu_j, \quad \nu_j \in [1/2, 2]$ & The site/instrument uncertainty scale factor. \\
\cutinhead{Keplerian Parameters}
$K$ & $1/K, \quad K \in \left[ \min_j \sigma_{y_j}/100, \max_j 10 \sigma_{y_j} \right]$ & RV semi-amplitude. \\
$P$ & $1/P, \quad P \in [600, 700] \, \mathrm{d}$ & Keplerian period \\
$e$ & $2e,\quad 0 \leq e < 1$\tablenotemark{b} & Keplerian eccentricity \\
$\omega$ & $\mathrm{const}, \quad 0 \leq \omega < 2\pi$ & Longitude of pericentre \\
$\chi$ & $\mathrm{const}, \quad 0 \leq \chi < 1$ & Pericentre passage is at $t = \chi P$\\
\cutinhead{CARMA Parameters}
$A_k$ & $1/A_k, \quad A_k \in \left[\min_j \sigma_{y_j}/100, \max_j 10 \sigma_{y_j} \right]$ & RMS amplitude of stochastic mode \\
$r_k$ & $1/r_k, \quad \Re r_k \in \left[ 1/ (2T), 0.05 \, \mu\mathrm{Hz}\right], \Im r_k = 0$\tablenotemark{c} & The damping rates for real modes \\
$f_k$ & $1/f_k, \quad f_k \in \left[ 0.1, 10 \right] \mu \mathrm{Hz}$ & The frequency of oscillatory modes \\
$Q_k$ & $1/Q_k, \quad Q_k \in \left[ 1, 1000 \right]$ & The quality factor of oscillatory modes \\
\enddata
\tablenotetext{a}{$\left\langle y_j \right\rangle$ and $\sigma_{y_j}$ are the mean and standard deviation of the RV timeseries from site and instrument combination $j$.}
\tablenotetext{b}{We actually sample in and impose a (flat) prior on $x \equiv e \cos \omega$ and $y \equiv e \sin \omega$.}
\tablenotetext{c}{Here $T$ is the time span of the measurements over all sites.}
\end{deluxetable}

\section{Aldebaran}
Aldebaran is a red giant star with spectral type K5, one of the nearest such stars at a distance of only $19.96 \pm 0.38$~pc as determined by \hipparcos \citep{hipparcos}. Its position near the Ecliptic permits the determination of its angular diameter by lunar occultations and by interferometry \citep[$20.58 \pm 0.03$ mas;][]{richichi2005,1979ApJ...228L.111B,brown1979,panek1980}. These tight constraints are valuable in breaking degeneracies in stellar modelling and make this an ideal star for asteroseismic characterization (Appendix~\ref{sec:mod}).

@@

\citet{Hatzes2015} reported on RV observations of Aldebaran from the coude
échelle spectrograph of the Thüringer Landessternwarte Tautenburg (TLS), the
Tull Spectrograph of the McDonald 2.7 m telescope, and the Bohyunsan Observatory
Échelle Spectrograph (BOES) spectrograph of the Bohyunsan Optical Astronomy
Observatory (BOAO) spanning the period from 2000.01 to 2013.92.  Combined with earlier observations \citep{hatzes1993}
